---
title: "Analysis of The Beatles Discography"
author: "Samuel Brown, Jay Chadha, Brett Carey"
date: "2022-12-11"
output: html_document
runtime: shiny
---

## Introduction

The Beatles were an English rock band formed in 1960 and were active until 1970. By the mid 1960’s they became an international sensation. With the rise of streaming services, older bands such as The Beatles uploaded their discography to platforms such as Spotify. Their music on Spotify was then transformed into a dataset, allowing for detailed analysis of each of their songs.

### Dataset and Variables
The dataset is comprised of 17 unique variables of that 7 were used:

- Album Name - Name of the Album
- Release Date - The date in which the album was released
- Popularity - A numeric value of the song’s popularity
- Danceability - A numeric value of the song’s danceability
- Energy - A numeric value of the song’s energy
- Tempo - A numeric value of the song’s tempo
- Loudness - A numeric value of the song’s loudness
- Valence -  A numeric value of the song’s valence

### Purpose
To determine if correlations or trends exist among different albums and its variables. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Read in libraries used
library(tidyverse)
library(shiny)
library(plotly)

# read in csv file
songs<- read.csv("beatles_spotify.csv")
```

## Question 1
Are there any trends over time for different song variables in the Beatles’ discography?

```{r, echo=FALSE}
# declare shiny app
shinyApp(
  ui = fluidPage(
  
  # title for shiny app
  titlePanel("The Beatles: Time vs Variable"),
  
  # sidebar for input variables
  sidebarLayout(
  sidebarPanel(
    
    # date selector for chooisng date inpu
    dateRangeInput("dates", label = h4("Date Range"), start=("1963-03-21")),

      # select input for choosing the color variable
    selectInput("selecty", label = h4("Choose Color Variable"), choices = list("Danceability" = "danceability", "Energy" = "energy", "Popularity" = "popularity","Loudness" = "loudness", "Tempo" = "tempo", "Valence"="valence"), selected = "danceability"),
  
    # select the variable being plotted
  radioButtons("radio", label=h4("Choose a Variable"), choices = list("Popularity"="popularity","Danceability"="danceability", "Energy"="energy","Tempo"="tempo", "Loudness"="loudness","Valence"="valence"), selected="popularity"),
  ),
  
    mainPanel(
    #plotly in main panel  
    plotlyOutput("boxplot"),
    hr(),
    fluidRow(column(3, verbatimTextOutput("value")))
    )
  )
  ),
  
  # call the server function
  server = function(input, output){
    
    # list of album that will be used to filter from the main dataset
    choicesList = list("1 (Remastered)","Let It Be (Remastered)","Abbey Road (Remastered)","Yellow Submarine (Remastered)","The Beatles (Remastered)","Magical Mystery Tour (Remastered)","Sgt. Pepper's Lonely Hearts Club Band (Remastered)","Revolver (Remastered)","Rubber Soul (Remastered)","Help! (Remastered)","Beatles For Sale (Remastered)","A Hard Day's Night (Remastered)","With The Beatles (Remastered)","Please Please Me (Remastered)")
    
    #output plot
    output$boxplot <- renderPlotly({
      ggplotly(
        # take main dataset and filter by the choices list above
        songs %>%
        filter(album %in% choicesList) %>%
          #filter again by if the release date is within the range from the input date range
          filter(release_date > input$dates[1] & release_date <  input$dates[2])%>%
          # ggplot call
        ggplot(aes(x = release_date, y = !!sym(input$radio), label = name)) +
          # change the point color based on the input variable to the dropdown
        geom_point(aes(color = !!sym(input$selecty)))+
          # theme ui and formatting
          theme_bw()+
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank())
        )
    })
  }, options = list(height = 550)
  
)
```

### Discussion 

This chart allows the user to select a date range for the songs to include in the chart and two song variables to visualize how they change over time. One variable will be plotted as the y variable and the other is encoded by varying the color of the points.

After selecting various variables to display over time, it does not seem that any clear trends appear in the Beatles’ discography over time. However, there are some interesting outliers. For one, the album Yellow Submarine, released in January 1969, appears to contain some of the Beatles’ least popular songs. This makes sense because this album contains the soundtrack to its namesake film, so the music is categorically different from their pop albums and therefore less popular. Another interesting outlier can be found by plotting the loudness over time, which reveals that the most recent album in the dataset, “1”,  has the loudest songs. “1” is a compilation album released in 2000, so improved production technology might be the reason for its songs’ higher loudness.



## Question 2
Are there any trends between different song variables in the Beatles' overall discography? 

```{r, echo=FALSE}
# call the shiny app
shinyApp(
  ui = fluidPage(
  titlePanel("The Beatles: Variable X vs Variable Y Based on Album"),
  
  # sidebar layout calls
  sidebarLayout(
  sidebarPanel(
    
    # Dropdowns to select x and y variables
  selectInput("selectx", label = h4("Variable X"), choices = list("Danceability" = "danceability", "Energy" = "energy", "Popularity" = "popularity","Loudness" = "loudness", "Tempo" = "tempo"), selected = "danceability"),
  
  selectInput("selecty", label = h4("Variable Y"), choices = list("Danceability" = "danceability", "Energy" = "energy", "Popularity" = "popularity","Loudness" = "loudness", "Tempo" = "tempo"), selected = "tempo"),
  
  # Slider to select span of loess line
  sliderInput("span", label = h4("Span:"),
              min = 0.2, max = 0.9, value = 0.5, step = 0.1),
  
    # This allows the user to check any album on Spotify
    checkboxGroupInput("checkGroup", label = h4("The Beatles Albums"), 
      choices = list("1" = "1 (Remastered)",
                     "Let It Be" = "Let It Be (Remastered)",
                     "Abbey Road" = "Abbey Road (Remastered)",
                     "Yellow Submarine" = "Yellow Submarine (Remastered)",
                     "The Beatles" = "The Beatles (Remastered)",
                     "Magical Mystery Tour" = "Magical Mystery Tour (Remastered)",
                     "Sgt. Pepper's Lonely Hearts Club Band" = "Sgt. Pepper's Lonely Hearts Club Band (Remastered)",
                     "Revolver" = "Revolver (Remastered)",
                     "Rubber Soul" = "Rubber Soul (Remastered)",
                     "Help!" = "Help! (Remastered)",
                     "Beatles For Sale" = "Beatles For Sale (Remastered)",
                     "A Hard Day's Night" = "A Hard Day's Night (Remastered)",
                     "With The Beatles" = "With The Beatles (Remastered)",
                     "Please Please Me" = "Please Please Me (Remastered)"), 
      selected = "Abbey Road (Remastered)"),
  ),
    
  # call the main panel and add in the plot
    mainPanel(
    plotlyOutput("scatter"),
    hr(),
    fluidRow(column(3, verbatimTextOutput("value")))
    )
  )
  ),
  
  # call the server
  server = function(input, output){
    
    # Make in the reactive scatterplot
    output$scatter <- renderPlotly({
      # ggplot call
      ggplotly(
        # filter by albums selected in the checkbox area
        songs %>%
        filter(album %in% input$checkGroup) %>%
          # plot the points based on the x and y variable selectged
        ggplot(aes(x = !! sym(input$selectx), y = !! sym(input$selecty), label = name)) +
          # color, theme, and formatting options
        geom_point(aes(color = album)) +
        theme_bw() +
        theme(legend.position = 'none') +
          geom_smooth(se = F, method = "loess", formula = y ~ x, span = input$span))
    })
  } , options = list(height = 900)
)
```

### Discussion

This chart lets the user select two variables and any amount of albums to plot songs by, creating a loess line of a selected span to visualize the relationship between the two variables. 
	
When all of the Beatles songs are plotted, there doesn't seem to be a strong trend between any two variables in a particular direction except energy & loudness. While the lack of relationships seems a little bit curious at first glance, the Beatles produced enough music in enough varying styles that their songs have covered practically every combination of musical variables around. It also makes sense that energy and loudness are correlated, as people are generally louder when they get excited.  
     